Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences

### 2017-08-15 paper-amia-summit-2018/main.tex
added initial files for AMIA Summit 2018

If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language. Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias. Figure~\ref{fig:delta} illustrates the accuracy of Markov Chain-based prediction method by varying the value of $\delta$. As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias. However, F-score sharply decreases after $\delta = -0.9$. Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/delta.png}
    \caption{\textbf{Accuracy of the proposed Markov model-based method by varying the value of $\delta$}.}
    \label{fig:delta}
\end{figure}

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, F-measure and accuracy using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). Accuracy of a class is defined as the ratio of the numbers of correctly classified sequences and a total number of sequences for that particular class in the gold standard.

@@

\subsection*{\textit{Predictive performance for natural PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449) and accuracy (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall and accuracy than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
 & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989} & 0.8059\\ \cline{2-6}
 & Second order & 0.7400 & \textbf{0.8449} & 0.7822  & \textbf{0.8449}\\ \hline

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353 & 0.9392\\ \cline{2-6}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
  \end{tabular}
\end{table}

>>>>

If $p(S\rightarrow CML+) > \delta $, where $\delta$ is an empirically determined threshold, the interview is predicted to result in positive change talk or commitment language. Tuning the proposed method on the training data involves experimentally determining the value of $\delta$, which results in the highest prediction F-measure with lower bias. Figure~\ref{fig:delta} illustrates the performance of Markov Chain-based prediction method by varying the value of $\delta$. As can be seen from Figure~\ref{fig:delta}, F-measure varies insignificantly, when $\delta $ is between -5 to -1, due to possible bias. However, F-score sharply decreases after $\delta = -0.9$. Thus, -0.9 is the optimal value of $\delta $, where F-measure is the highest with lower bias.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/delta.png}
    \caption{\textbf{F-Measure of the proposed Markov model-based method by varying the value of $\delta$}.}
    \label{fig:delta}
\end{figure}

@@

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 5-fold cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)).

@@

\subsection*{\textit{Predictive performance for natural PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532\\\cline{2-5}
 & Second order & 0.6889 & 0.7889 & 0.7352\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989}\\ \cline{2-5}
 & Second order & 0.7400 & \textbf{0.8449} & 0.7822\\ \hline

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604\\\cline{2-5}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736}\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353\\ \cline{2-5}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699\\ \hline
  \end{tabular}
\end{table}
